Are we learning to build better models at the cost of better decisions?

### 2021-03-06 manuscript/manuscript.Rmd
:pen:

Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management. One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.  While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty [@Walters1981; @Smith1981].  In this approach, a manager assigns probabilities to each of the possible models under consideration, i.e. models 1 & 2.  The expected utility of any action reflects both the intrinsic uncertainty over future states under either model (process uncertainty) and the manager's uncertainty over the choice of models (or parameter values). Unlike the forecast comparison, an adaptive management approach explicitly considers the utility function, Eqn \eqref{utility}, which it seeks to maximize by intergrating over model uncertainty.

In passive adaptive management, the manager chooses the action which maximizes that expected future utility. After observing the consequences of the action, the manager updates the posterior probabilities that each model is correct, typically by Bayes rule [e.g. @Walters1981; @Smith1981; @Ludwig1982; @Punt2016]. The decision strategy is then re-evaluated in light of this revised uncertainty.  In many real-world applications, it may be impossible or computationally prohibitive to solve the optimal control policy given all sources of uncertainty, as is typical in fisheries management practice [@Punt2016]. Management Strategy Evaluation (MSE) evaluates

In active adaptive management, the key difference is that the manager may chose an action which does not maximize the utility _given the current model uncertainty_, but instead, takes an action whose consequences will more quickly discriminate between the models and thus more rapidly reduce the uncertainty. After observing the outcome of this exploratory step, model probabilities are updated as before.  Under some scenarios the cost of this 'exploration' step can be made up later by actions based on the more precise knowledge gained as a result ('exploitation'). In practice, passive adaptive management approaches are more common, such as Management Strategy Evaluation (MSE @Punt2016) or Scenario Analysis [@Polasky2011]. These approaches all share the same adaptive aspect,

Owing to the computational complexity involved, these approaches typically evaluate a suite of proposed policies, rather than seeking to derive the optimal policy. A scenario analysis does not rely on quantitative estimates of model probabilities or expected utility, but still consider possible outcomes over a suite of models (scenarios), select actions to minimize the risk or or other suitable objective, and update or eliminate scenarios deemed implausible over time [@Polasky2011].

Each of these approaches swiftly decides that model 1 in our example is implausible, and settles for making decisions driven almost exclusively by model 2.  I illustrate this using a passive adaptive management strategy following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty. To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.  After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig 4]. As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig 2, achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone]. If instead of considering only these two models we consider a whole suite of Gordon-Schaefer-type models with varying $r$ and $K$ parameters, learning is slightly slower but no less counter-productive: mean stock size is $46%$ and net present value is $20%$ of what would be achieved under Model 1 (Appendix, Fig S1).

@@

I think it would be effec

@@

@@

Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other.  The reason for this may appear obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. Nevertheless, decisions based on model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.  So why do our best approaches fail such a simple test case? It is tempting to dismiss this example as having little relevance to real-world decision problems on a variety of grounds: (1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of an unrealistic optimal control solution, (3) the problem arises from having started with different objectives, (4) the problem is already solved. I address these in turn.

@@

>>>>

Decision-theoretic approaches include scenario analysis, resilience thinking, optimal control and related methods [@Polasky2011], which have a long history in ecology and particularly in fisheries management. One of the most influential of these is _adaptive management_ [sensu @Walters1978], in which a manager seeks to both reduce uncertainty over time while also achieving the best outcomes given current knowledge.  While the term today is frequently used in a looser sense, adaptive management as originally developed can be considered an example of optimal control under model uncertainty [@Walters1981; @Smith1981].  In this approach, a manager assigns probabilities to each of the possible models under consideration, i.e. models 1 & 2.  The expected utility of any action reflects both the intrinsic uncertainty over future states under either model (process uncertainty) and the manager's uncertainty over the choice of models (or parameter values). Unlike the forecast comparison, an adaptive management approach explicitly considers the utility function, Eqn \eqref{utility}, which it seeks to maximize by integrating over model uncertainty. After observing the consequences of the action, the manager updates the posterior probabilities that each model is correct, typically by Bayes rule [e.g. @Walters1981; @Smith1981; @Ludwig1982; @Punt2016]. The decision strategy is then re-evaluated in light of this revised uncertainty.  Adaptive management may be either passive or active. Under passive adaptive management, the decision-maker chooses the action which maximizes that expected future utility.  Under active adaptive management, the decision-maker may chose 'exploratory' actions which are not optimal under the current uncertainty but may lead to more rapidly decreasing the uncertainty in the future [@Walters1978]. In practice, many managers are reluctant to employ exploratory strategies, so passive adaptive management approaches such as as Management Strategy Evaluation (MSE) are more common [@Punt2016].

<!--
Any of these adaptive management approaches proves to be maladaptive under the simple example of model uncertainty considered here.  That is, the manager derives lower utility (here, economic yield), and the fish stock is also surpressed to a lower biomass than would be achieved if an identical management approach without any adaptive updating to model uncertainty were employed.  Put another way, the "value of information" gained from future observations of stock in response to any management action is actually negative.

Owing to the greater complexity involved in many fisheries models, a typical MSE would evaluate a fixed set of possible strategies rather than solve a sequential decision problem [@Marescot2013] for the optimal sequence of actions, but the central premise of iteratively updating model posterior probabilities remains the same.  By limiting ourselves to the simpler models considered here, we are able to perform a passive adaptive management analysis of the full sequential decision problem, analgous to earlier examples [@Walters1981; @Smith1981; @Ludwig1982], which ensures the result is not an artifact of considering too narrow a set of possible policies (e.g. only constant-mortality policies).

Unfortunately, any adaptive management strategy which updates posterior distributions that characterize model undertainty leads to worse outcomes than an approach which does no updating.  I illustrate the application of a passive adaptive management strategy to this simple example, following classic examples for parameter [@Ludwig1982] or structural [@Smith1981] model uncertainty.  To demonstrate that the behavior is not driven by failure to explore sufficiently, (which might be addressed by an active adaptive management), I will assign initial probability that model 2 is true at 1%.  After a single iteration of learning, these probabilities are completely reversed, with the manager deciding that model 2 is almost certainly correct  [Fig 4]. As before, this results in a management practice with much worse ecological and economic outcomes than would have been realized by a manager who stubbornly clung to model 1 without updating [Fig 2, achieving a net present value over 100 time steps that only 31% that expected under management using Model 1 alone]. If instead of considering only these two models we consider a whole suite of Gordon-Schaefer-type models with varying $r$ and $K$ parameters, learning is slightly slower but no less counter-productive: mean stock size is $46%$ and net present value is $20%$ of what would be achieved under Model 1 (Appendix, Fig S1).

@@

@@

- is this the result of underestimating the uncertainty?

@@

Given this simplified version of model uncertainty in which one of the two models leads to effectively optimal ecological and economic outcomes, current approaches invariably choose the other. Importantly, a decision maker employing an adaptive management or iterative forecasting assessment such as those considered here would have no way of realizing that the outcomes they experienced were in fact sub-optimal.  The manager quickly concludes that Model 1 is entirely implausible, but also finds Model 2 to be remarkably accurate at predicting future values, and is entirely unaware that the model is leading to such poor outcomes. A more diligent manager may seek to improve Model 2 further by re-estimating parameters, but this in fact leads to a model with slightly lower $K$ and thus even higher level of over-fishing and under-performance than Model 2. The only way to avoid this trap is not to learn with each new observation, not to apply iterative updates over the models. A manager must follow the decisions proposed by model 1 for many iterations to realize that it leads to more desirable outcomes.  This raises several questions.  Do such methods of learning and reducing model uncertainty ever lead us towards worse long-term outcomes in the real world? How would we know? And what should we do about it?

It is tempting to argue that this situation would almost never arise in real world scenarios.  In this argument, Model 1 is a unicorn, a mythical model that somehow gives optimal performance while at the same time making predictions that are wildly wrong.  Could such unicorns really exist?  This is a difficult assertion to prove either way, but it is worth noting that neither of these features by themselves are especially rare. Examples of serious forecasts that widely miss future observations abound wherever forecasts are common, from elections to economics to environmental change [@].  Meanwhile, the premise that model need not perfectly capture reality to be useful is at the very heart of modeling.  Model 1 successfully captures the one key feature driving decisions in optimal harvest control problems: the stock size at which the maximum growth rate occurs [Fig 4a].  Capturing only the essential aspects as simply as possible is the goal of any model building exercise.  Moreover, rigorous updating of models and parameters through approaches such as adaptive management or iterative forecasting are not yet widespread practices in ecological management [@Dietze2018].  If such unicorn models exist, that lack of adoption would make it easier for them to persist at least until now.

If these unicorns do exist in real world management, then what do we do about it? Our step may be to acknowledging that we do not yet have the answer.  There are at least three common ideas which are not the answer, which might succinctly be termed (1) don't care, (2) don't learn or (3) don't model.

The first response, don't care, suggests that iterative updating is always desirable.  Previous work has acknowledged that future observations may be relatively uninformative, failing to reduce uncertainty between models by much.  This is the motivation for active over passive adaptive management, [@Walters1978], and has also acknowledge that even significant improvements in predictive accuracy may have negligible improvement on the quality of the decision.  This is the motivation for quantifying the value of information, [@Katz1987].  However, the conclusion from such observations is that additional information is at worst harmless to the quality of the decision, while in this example it is actively detrimental.  The realized 'value of information' contained in future measurements is negative.

The

Why then do our best available approaches so readily reject the model?

The reason for this may appear obvious once we compare both curves to that of the underlying model, Model 3.  Plotting the growth rate functions of each model, [Fig 4a], it is hardly surprising that no method which would prefer the closely overlapping Model 2 to the no-where-close Model 1 as the better approximation of Model 3. Nevertheless, decisions based on Model 1 are nearly indistinguishable from those based on the true model [Fig 4b], while Model 2 leads to over-harvesting.  We have no methods of reducing model uncertainty which do not perform significantly worse on this management problem than an approach which does not attempt to learn.  Put another way, the value of information contained in future observations [@] is always negative in this example.

So why do our best approaches fail such a simple test case? It is tempting to dismiss this example as having little relevance to real-world decision problems on a variety of grounds:   (3) Don't we already know this?

(1) this problem is an artifact of overly-simplified models, (2) this problem is an artifact of an unrealistic optimal control solution, (3) the problem arises from having started with different objectives, (4) the problem is already solved. I address these in turn.

@@

This issue is not unique to sequential decision problems or optimal control solutions.  For example, Management Strategy Evaluation [@Punt2016] in fisheries seeks to evaluate pre-specified strategies rather solve for the best possible strategy through optimal control.  This approach is well justified -- optimal control techniques such as stochastic dynamic programming illustrated here quickly become intractable for more complex models [@Marescot2013] typically used in fisheries stock assessments.  Constraining the search to only strategies that impose a constant mortality (defined as harvest per unit biomass, $F = H/B$) means we only have to evaluate those $N$ strategies each iteration, not solve a sequential decision problem. Doing so, we would find the best constant-mortality solution under Model 1 performs much worse than the best constant-mortality solution under Model 2. Does this mean MSE is not suseptible to this issue?  No, it does not.  It is just as easy to construct an alternative Model 1 with the same properties of leading to nearly optimal decisions while being rejected by any method of iterative learning, such as forecasting or adaptive management.  Just as the optimal control policy for a Gordon-Schaefer (logistic) model depends only on parameter $K$ [@Reed1979], the optimum constant escapement policy depends only on parameter $r$ [@Schaefer1954]. If Model 1 has a value of $r$ such that it happens to match the best possible constant-mortality solution for the (unobserved) true model, then $K$ can be set arbitrarily high, ensuring the any model selection, forecasting, or adaptive management approach would lead away from Model 1 and towards worse-performing models once again.
